---
title: "FDR_lasso"
author: "Mathieu Guglielmino"
date: "21/01/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('TPP_FDP.R')
source('sharpness.R')
```

## Lasso path

We generate data under a random gaussian design, with $n = 1010$ and $p = 1000$. We chose $\beta$ such that $ \beta_1 = \cdots = \beta_200 = 4$ and $\beta_201 = \cdots = \beta_800 = 0$.

```{r data_generation}
beta = c(rep(4, 200), rep(0, 800))

X = matrix( rnorm(1010 * 1000), nrow=1010, ncol=1000)
y = X%*%beta
```

We plot the lasso path for 500 values of $\lambda$, for $\lambda$ varying between 0.1 and 10.

```{r lasso_path}
lasso_path <- t(sapply(seq(0.1, 10, length.out = 500), function(x) TPP_FDP(x, X, y, beta) ))
plot(lasso_path,
     xlab = 'TPP', ylab = 'FDP', main = 'Lasso path', cex=0.5, col="deepskyblue3")
```
### Comments
How to chose the values of $\lambda$ ?

## TPP at time of first false selection and FDP at time of last true selection
We generate 100 datasets under the previous conditions for which we would like to know the frequency of both : 
+ The TPP at time of first false selection
+ The FDP at time of last true selection

Let's recall that for $\lambda = 0$, we are under the least squares estimate framework, hence every variables will be selected. Under this condition the TPP will equal one, and we increase $\lambda$ to figure the

```{r tpp_fdp_histogram}
# Histogram of the TPP at time of first false selection
tpp_first_false_selection <- sapply(1:100, function(x) first_false_selection(1010, 1000, beta)[1])
# Histogram of the FDP at time of last true selection
fdp_last_true_selection <- sapply(1:100, function(x) last_true_selection(1010, 1000, beta)[2])

par(mfrow=c(1,2))
hist(tpp_first_false_selection, col = "gold2",
     xlab = "TPP at time of first false selection", main="")
hist(fdp_last_true_selection, col = "gold2",
     xlab = "FDP at time of last true selection", main="")

```

On left, the TPP when the first false variable enters the lasso. Right, the FDP when the TPP reaches one.

## Sharpness

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
